---
date: "2019-09-03T10:14:46+02:00"
title: Importing Data
weight: 12
---


## Importing data and the `here()` function  

**The templates use the function `here()` in conjunction with `import()` to locate and import local files.** 

### The `import()` function  

The function `import()` from the package **rio** makes it easy to import many types of files.  

It uses the file's extension (e.g. .xlsx, .csv, .dta, etc.) to appropriately read the file. You can read more about `import()` [on this page](https://cran.r-project.org/web/packages/rio/vignettes/rio.html). If importing an Excel file, the templates use `import()`'s argument `which=` to specify the sheet to import.  

However, **`import()` needs to be told where to look on your computer** to find the file. **Therefore, we have wrapped `import()` around the function `here()`**. Normally, `import()` accepts a filepath in quotes, but it is easy to make mistakes that way. 

### The `here()` function  

The function `here()` from the package **Here** makes it easy to locate local files.  

`here()` works well within R projects. When the package is first loaded, it automatically considers the top-level folder of your R project as "here" - as a benchmark for all other files in the Project. In your script, if you want to import or reference a file saved in your R project's folders, you use the function `here()` to tell R where the file is in relation to that benchmark.  

If you are unsure where "here" is set to, run this command:

```{r eval=FALSE}
# This command tells you the folder path that "here" is set to 
here()
```

### Importing using `import()` and `here()` together  

**Below is an example** of importing the file "linelist.xlsx". If the file you want is in the folder designated as "here", all you have to do is provide the name of the file in quotes (with the appropriate ending). 

```{r eval=FALSE}
linelist_raw <- rio::import(here("linelist.xlsx"), which = "Sheet1")
```

{{% notice tip %}}
After the closing parenthes of `here()` the `import()` function continues. There is a comma and `import()`'s optional argument `which = `, specifying which sheet of the Excel file to import.
{{% /notice %}}

**If however, your file is within subfolders of the folder designated as "here"** - let's say it is in the subfolder "data" and then the sub-subfolder "linelists" - **you write these folder names before the file name, in quotes, and separated by commas**, as below:  

```{r eval=FALSE}
linelist_raw <- rio::import(here("data", "linelists", "linelist.xlsx"), which = "Sheet1")
```

You can read more about Here and `here()` in [this overview](https://github.com/jennybc/here_here) and [this documentation](https://here.r-lib.org/)  

{{% notice tip %}}
By using the **Here** package and `here()` function with R projects, it is not necessary to worry about your "working directory", `setwd()`, or `getwd()` commands.
{{% /notice %}}

