---
date: "2019-08-21T10:32:03+02:00"
title: Standardise and clean data
menuTitle: Clean data
weight: 11
---


```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# hide all code chunks in the output, but show errors
knitr::opts_chunk$set(echo = TRUE,       # hide all code chunks in output
                      error = TRUE,       # show errors if they appear, but don't stop
                      fig.width = 6*1.25, # Figure width
                      fig.height = 6      # Figure height
                     )
# set default NA to - in output, define figure width/height
options(knitr.kable.NA = "-")

# Installing required packages for this template
required_packages <- c("knitr",       # create output docs
                       "dplyr",       # clean/shape data
                       "forcats",     # clean/shape data
                       "stringr",     # clean text
                       "rio",         # read in data
                       "ggplot2",     # create plots and charts
                       "sitrep",      # MSF field epi functions
                       "linelist",    # Functions for cleaning/standardising data
                       "incidence",   # create epicurves
                       "aweek",       # define epi weeks
                       "epitools",    # 2x2 tables and other epi goodness
                       "epitrix",     # epi helpers and tricks
                       "sf",          # encode spatial vector data
                       "ggspatial",
                       "here")   # plot maps

for (pkg in required_packages) {
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  
  # load packages to this current session 
  library(pkg, character.only = TRUE)
}


# set default text size to 16 for plots
# give classic black/white axes for plots
ggplot2::theme_set(theme_classic(base_size = 18))

# Set the day that defines the beginning of your epiweek.
# you can start the week on any day of the week
# (the ISO standard is to start on Monday) 
aweek::set_week_start("Monday")



date2week("2017-04-20")      #Sys.Date() uses the current date from your computer


reporting_week <- "2017-W16"  # Set the reporting week

# Read in the R object linelist_cleaned that is defined and saved 
# in the AJS_Walkthrough_Cleaning script
linelist_cleaned <- readRDS("data/linelist_aligned.rds")


```




## Standardize and clean data  
Becuase the Am Timan dataset is not yet aligned with the MSF data dictionary and what the template expects, there are several data cleaning steps we must complete.  

### Re-code missing values from periods (.) to `NA` at import  

Currently, missing values in the dataset are represented by a period (.). This causes problems because R expects missing values to be coded as `NA`. For example, if we attempt to set the variable `date_of_onset` to class Date, R gives an error because the periods are considered character values that do not fit the expected YYYY-MM-DD date format.  

To fix this problem we **scroll up to the `read_nonDHIS_data` chunk of the template and insert an additional argument in the `rio::import()` command**. The argument `na = "."` specifies a period as the value in the Excel sheet that R will consider to be "missing". As the data are re-imported, all cells with that value are now changed to `NA`.  

**Remember that after making this change to the import command you must re-run the entire script. This way, the dataset is correctly imported AND the later cleaning steps, changes to variable names, etc. are also performed on this newly-imported dataset** 

```{r eval=FALSE}
# The old import command
linelist_raw <- rio::import("AJS_AmTiman.xlsx", which = "linelist")

# The revised import command (don't forget the comma between arguments!)
linelist_raw <- rio::import("AJS_AmTiman.xlsx", which = "linelist", na = ".")
```

### Drop ineligible observations  
The next code item in the script drops observations with missing `case_number` or `date_of_consultation_admission`. Because the Am Timan dataset contains observations for patients seen in the community and at the hospital, these criteria may not be relevant. We can check and see that of the 1447 observations, there are 0 missing `case_number` or `dateentry`, but 616 observations missing `date_of_consultation_admission`. These 616 community-identified cases are still of interest for our report, so we will not drop them. 

```{r eval=FALSE}
# Find the number of observations missing date_of_consultation_admission
nrow(linelist_cleaned %>% filter(is.na(date_of_consultation_admission))) 
```


{{% notice tip %}}
Notice that the command above returns the number of rows missing `date_of_consultation_admission`, but does not overwrite `linelist_cleaned` with an assignment operator `<-`. 
{{% /notice %}}


### Convert and clean date variables
In linelist_cleaned there are seven variables containing dates (`dateentry`, `epiquestdate`, `date_of_consultation_admission`, `date_of_onset`, `date_lab_sample_taken`, `dthospitalisation`, `dtdeath`).  

Because the Am Timan dataset contains date variables not found in the data dictionary, we will ignore and comment out the first code of the Date variables chunk (these commands convert only the date variables known to the data dictionary).  

Instead, we uncomment and **use the commands applicable to when a dataset does not align with the data dictionary**. You may see messages warning that a few dates are not in the correct timeframe - this is ok and these dates will be addressed in the next step.

```{r results='hide', message=FALSE, warning=FALSE}
# Use this is your dataset aligns with the data dictionary
# Creates list of date variables from the data dictionary
#DATEVARS <- filter(linelist_dict, data_element_valuetype == "DATE") %>% 
#  select(data_element_shortname) # select date vars
# change to the date variables found in your dataset to class date 
#linelist_cleaned <- linelist_cleaned %>%
#  mutate_at(DATEVARS$data_element_shortname, as.Date)

# Use this if your dataset does not align with the data dictionary
  linelist_cleaned <- linelist_cleaned %>%
    mutate_at(vars(matches("date|Date")), linelist::guess_dates,
            error_tolerance = 0.5)
```

The code above converted successfully many of our date variables (remember you can check class like `class(linelist$date_of_onset`). However, did not detect the variables named with "dt", such as `dtdeath` and `dthospitalization`. We must **convert those variables manually**:  

```{r}
# Individually convert other date columns
class(linelist_cleaned$dtdeath)

linelist_cleaned$dtdeath <- linelist::guess_dates(linelist_cleaned$dtdeath)

class(linelist_cleaned$dtdeath)

# Repeat for variable "dtdeath"
linelist_cleaned$dthospitalisation <- linelist::guess_dates(linelist_cleaned$dthospitalisation)
```

Next we uncomment code to **correct unrealistic dates**. We have browsed our data and know that there are observations with `date_of_onset` outside the reasonable range:
```{r}
# Check range of date_of_onset values, ignoring (removing) missing values
range(linelist_cleaned$date_of_onset, na.rm = TRUE)
```

We convert dates outside the expected range (April 2016 to October 2017) to missing using `case_when()`. Note that when making the assignment on the right-hand side (RHS), wrap NA in `as.Date()`.  

As you modify this chunk for your own situation, remember to include commas at the end of each `case_when()` line (except the last one), and that **each left-hand side (LHS) of the `~` must be a logical statement**, not just a value. Best is to write `case_when()` lines from most specific at top to most general at bottom. You can read more about `case_when()` [on this tidyverse site](https://dplyr.tidyverse.org/reference/case_when.html).  


{{% notice tip %}}
All the right-hand side (RHS) values must be the same class (either character, numeric, etc.). So, if your other RHS values are character and you want one RHS value to be missing, you cannot just write `NA` on the RHS. Instead you must use the special character version of `NA` : `NA_character_`.
{{% /notice %}}

  

```{r}
# Convert dates before April 2016 or after October 2017 to missing (NA)
linelist_cleaned <- mutate(linelist_cleaned,
                           date_of_onset = case_when(
                                 date_of_onset < as.Date("2016-04-01")  ~ as.Date(NA),
                                 date_of_onset > as.Date("2017-10-31")  ~ as.Date(NA),
                                 TRUE                                   ~ date_of_onset
                                 ))
                           
```

We also must **use the provided code to create a variable called `epiweek`**. While there are already variables in our dataset that give the epidemiological weeks of various data points, it is safer to make a new variable AND later code chunks rely on the variable being named `epiweek`. See the [documentation of the package aweek](https://cran.r-project.org/web/packages/aweek/vignettes/introduction.html) for more information.

```{r}
# Create variable epiweek
linelist_cleaned$epiweek <- aweek::date2week(linelist_cleaned$date_of_onset, 
                                             floor_day = TRUE, 
                                             factor = TRUE)
```



{{% notice tip %}}
A note on the epiweek variable and factors:  
Notice that we converted dates to NA that were clearly errors **before** creating the `epiweek` variable. This is because the `epiweek` variable is created as a **factor**, and once a factor is created it is more difficult to remove factor levels. YOu can read more about factors and dropping unused factor levels in the Advanced R Basics page.* 
{{% /notice %}}



### Clean numeric variables  

We can **comment out/ignore the code that makes a new numeric variable `obs_days`**, because we do not have a `date_of_exit` in our dataset.  

**However, we do need to clean the variable `age_years`**. If we look at the range of values, we get something strange:  
```{r}
# See the range of age_years values, removing (excluding) NA
range(linelist_cleaned$age_years, na.rm = TRUE)
```
We know there are ages older than 9 years. We check `class(linelist_cleaned$age_years)` and see that R is reading this variable as class *character*, not numeric! So we must convert it by adding the following command to the script:
```{r}
# Convert variable age_years to numeric class
linelist_cleaned$age_years <- as.numeric(linelist_cleaned$age_years)
```

**Now we must further clean the age-related variables.** In the chunk "Age group variables", we do not need and can comment out the code to add under 2 years to the `age_years` variable. However, we do have some decimal years. To align with the data dictionary we uncomment and **use the code to create a *new* variable `age_months` from the decimal years**.  

```{r}
# For patients under 5 years, the new variable age_months is based on age_years * 12 
linelist_cleaned <- mutate(linelist_cleaned,
                            age_months = case_when(
                              age_years < 5 ~ age_years * 12
                              )) 
```

We can then use the code to create an age group variable based on `age_months`.  

```{r}
## create age group variable for under 5 years based on months
linelist_cleaned$age_group_mon <- age_categories(linelist_cleaned$age_months, 
                                                 breakers = c(0, 6, 9, 12, 24), 
                                                 ceiling = TRUE)
```

...and use the similar code for groups of `age_years`. A note on grouping using `age_categories()`: for example with breakers 0, 3, 15, 30, and 45, a 30-year old patient will be included in an age_group named "30-44".

```{r}
## create an age group variable by specifying categorical breaks
linelist_cleaned$age_group <- age_categories(linelist_cleaned$age_years, 
                                             breakers = c(0, 3, 15, 30, 45))
```

### Clean categorical variables  
While the template directs us to next create a `DIED` variable based on `exit_status` containing the characters "Dead", we must first **clean our `exit_status` variable, which is currrently in French**. 

**Use `case_when()` to assign new values for a new exit_status2 variable**, as in the code below. The code uses the function `case_when()` to re-define `linelist_cleaned` as itself, but mutated to create the new variable `exit_status2`. The values in `exit_status2` are based on the values in `exit_status`, such that when `exist_status == "Décédé"`, the value in `exit_status2` is "Died", and so on. 

As you modify this chunk for your own situation, remember to include commas at the end of each `case_when()` line (except the last one), and that **each left-hand side (LHS) of the `~` must be a logical statement**, not just a value. Best is to write `case_when()` lines from most specific at top to most general at bottom. You can read more about `case_when()` [at this tidyverse site](https://dplyr.tidyverse.org/reference/case_when.html).  

{{% notice tip %}}
All the right-hand side (RHS) values must be the same class (either character, numeric, etc.). So, if your other RHS values are character and you want one RHS value to be missing, you cannot just write `NA` on the RHS. Instead you must use the special character version of `NA` : `NA_character_`.{{% /notice %}}


```{r}
# Tabulate to see all possible values of exit_status
table(linelist_cleaned$exit_status, useNA = "always")

# Create exit_status2 from values in exit_status
linelist_cleaned <- linelist_cleaned %>% 
  mutate(exit_status2 = case_when(
    exit_status == "Décédé"          ~ "Died",
    exit_status == "-"               ~ NA_character_,
    exit_status == "Echappé"         ~ "Left",
    exit_status == "Déchargé/Guéri"  ~ "Discharged"
  ))

# Tabulate the NEW exit_status2 variable to check correct assignment
table(linelist_cleaned$exit_status2, useNA = "always")
```


Now we can **make the `DIED` variable**, which is referenced in later code chunks. This command creates `DIED` as a logical (`TRUE` or `FALSE`) variable, depending on whether each observation meets the criteria to the right of the assignment operator `<-`. For a similar example, read about the `%in%` operator on the Advanced R Basics page.  

```{r}
## Note we are directing R to look within the NEW exit_status2 variable
linelist_cleaned$DIED <- linelist_cleaned$exit_status2 == "Died"
```


### **Re-code values in `patient_facility_type`**  

When we assigned our variables to match the data dictionary, we used the variable `hospitalisation` as the variable `patient_facility_type`. **However, the values in that variable do not match those expected by the template.** In the data dictionary, `patient_facility_type` should have values of "Inpatient" or "Outpatient." Currently, the values are "Oui" and "Non". In later code chunks, analyses are restricted to observations where `patient_facility_type == "Inpatient"`, thus, we should align the values to match the data dictionary.  

**To clean these values we uncomment and modify code from the `standardise_clean_data` chunk, found under the heading "recode a character variable".** 

```{r}
# View all the values in patient_facility_type
table(linelist_cleaned$patient_facility_type, useNA = "always")

# Convert the values
linelist_cleaned <- linelist_cleaned %>% 
  mutate(patient_facility_type = case_when(
    patient_facility_type == "Oui" ~ "Inpatient",
    patient_facility_type == "Non" ~ "Outpatient"
  ))

# Re-check that the values converted sucessfully
table(linelist_cleaned$patient_facility_type, useNA = "always")

```

### Recode values in lab test variables  

For later steps, we will need many of our testing variables to the values "Positive" and "Negative" instead of "Pos" and "Neg". We can uncomment and use this helper code to make that change. Some notes about this step:  

* You may receive warnings about unknown levels in some variables - this is generally okay but check the changes visually to be sure `View(linelist_cleaned)`.  
* One variable (`test_hepatitis_e_igm`) has values 0 and 1 that we want to change to "Positive" and "Negative". The `fct_recode()` function expects *character* RHS inputs (within quotes) - so put "0" and "1" in quotes, as below, and confirm the change.  
* One variable (`test_hepatitis_b`) has a value recorded as "Weakly pos". For the purposes of this exercise we re-code this to "Positive".

```{r, warning=FALSE}
## sometimes, coding is inconsistent across variables -- for example, "Yes" / "No"
## may be coded as Y, y, yes, 1 / N, n, no, 0. You can change them all at once!
## Create a list of the variables you want to change, and run the following.
## You may need to edit this code if options are different in your data.

# # create list of variables
change_test_vars <- c("hep_e_rdt", "hep_c_rdt", "hep_b_rdt", "test_hepatitis_a", "test_hepatitis_b", "test_hepatitis_c", "malaria_rdt_at_admission", "test_hepatitis_e_genotype", "test_hepatitis_e_igm", "test_hepatitis_e_virus", "hevpcr", "other_arthropod_transmitted_virus")

# standardize options
  linelist_cleaned <- linelist_cleaned %>%  
       mutate_at(vars(change_test_vars), forcats::fct_recode,   
       Positive = "Pos",
       Positive = "pos",
       Positive = "yes",
       Positive = "Yes",
       Positive = "Weakly pos",
       Positive = "1",
       Negative = "Neg",
       Negative = "neg",
       Negative = "no",
       Negative = "No",
       Negative = "0"
     )  
``` 


### Other categorical variables  

We can **comment/ignore the code to change the order of levels in categorical variables**, as the Am Timan dataset does not include a variable `time_to_death` and we do not need to change the order of any categorical variables.  

**However, we do need to create a case definition variable.** In this use of `case_when()`, the last line left-hand side (LHS) is TRUE, which serves as a catch-all for any other possible values that have not yet met the criteria of the earlier `case_when()` lines.  

Note how this code checks the variable `hep_e_rdt` for "Positive". The earlier cleaning steps converting `hep_e_rdt` values from "Pos" to "Positive" were necessary for this case definition to properly apply.  

In addition we need to change the value looked for in `other_cases_in_hh` from "Yes" to "Oui"

```{r}
# You MUST modify this section to match your case definition. The below
# uses positive RDT for Confirmed and epi link only for Probable.

linelist_cleaned <- linelist_cleaned %>%
  mutate(case_def = case_when(
    is.na(hep_e_rdt) & is.na(other_cases_in_hh)           ~ NA_character_,
    hep_e_rdt == "Positive"                               ~ "Confirmed",
    hep_e_rdt != "Positive" & other_cases_in_hh == "Oui"  ~ "Probable",
    TRUE                                                  ~ "Suspected"
  ))
```

### Cleaning patient origin  

The variable `patient_origin` was originally the variable *quartier* in the raw dataset. **patient_origin is referenced in the place analyses chunks**, for example when calculating attack rates by region. In these steps, the variable `patient_origin` in `linelist_cleaned` is *joined to* the `patient_origin` variable of the data frame `population_data_region`, which was imported in the population and lab data steps.  

**However, the place names in `population_data_region` are ALL CAPITAL LETTERS and this is not true in `linelist_cleaned` - the join will not register any matches.** The easiest way to solve this problem is to redefine the linelist_cleaned variable into all capital letters as well, using the base R function `toupper()`:  

```{r}
linelist_cleaned$patient_origin <- toupper(linelist_cleaned$patient_origin)
```

### Create lists of variables  

These variables names are stored in vectors that are created using the function `c()`. These vectors of names will be referenced in later code chunks. This way if you want to run the same function over these variables you can simply use the named vector rather than typing out each variable individually.  

The default template code creates two vectors - one for symptoms variables (`SYMPTOMS`) and one for laboratory testing variables (`LABS`). However, *we can see that there are two rather different type of lab variables*: those with Positive/Negative values, and those with IgG and IgM results. To make our table of lab results more clean, we split the lab variables into these two groups.

**Some variables do not apply to our Am Timan data and must be removed from the list or commented out (such as `convulsions` and `history of fever`).** If non-existant variables are left in the list it will cause errors in later code chunks.

```{r}
# vectors of variable names ----------------------------------------------------
# create a grouping of all symptoms 

SYMPTOMS <- c("generalized_itch", 
#             "history_of_fever", 
              "fever",
              "joint_pains",
              "epigastric_pain_heartburn",
              "nausea_anorexia",
              "vomiting", 
              "diarrhoea",
              "bleeding", 
              "headache",
              "mental_state",
#              "convulsions",
              "other_symptoms"  
              )
```


```{r}
# create a grouping of all lab tests 
LABS <- c("hep_b_rdt", 
          "hep_c_rdt",
          "hep_e_rdt",
          "test_hepatitis_a",
          "test_hepatitis_b",
          "test_hepatitis_c",
#          "test_hepatitis_e_igg",
          "test_hepatitis_e_igm" ,
          "test_hepatitis_e_genotype",
          "test_hepatitis_e_virus",
          "malaria_rdt_at_admission",
#          "malaria_blood_film", 
          "dengue",
#          "dengue_rdt", 
          "yellow_fever",
#          "typhoid",
          "chikungunya_onyongnyong", 
#          "ebola_marburg",
#          "lassa_fever",
          "other_arthropod_transmitted_virus" 
#          "other_pathogen"
          )

```

### Dropping observations  

The next code assists with **dropping observations and variables**. We can remove any observations with date of onset after the `reporting_week` (`reporting_week` is an object defined at the beginning of the script).  


{{% notice tip %}}
The default code in the template also removes observations with missing (NA) `date_of_onset`. We can add ` | is.na(date_of_onset)` into the filter to keep these.
{{% /notice %}}

```{r}
# Drop cases after reporting week
linelist_cleaned <- linelist_cleaned %>% 
  filter(date_of_onset <= week2date(sprintf("%s-7", reporting_week)) | is.na(date_of_onset))
```

Then, it is important to **define these date-related objects** that are referenced in later code chunks.  

```{r}
# define the first week of outbreak (date of first case)
first_week <- levels(linelist_cleaned$epiweek)[1]

# outbreak start 
# return the first day in the week of first case 
obs_start <- week2date(sprintf("%s-1", first_week))

# return last day of reporting week 
obs_end   <- week2date(sprintf("%s-7", reporting_week))

```

### Export if desired  

And finally, if desired you can **export the cleaned dataset** for other purposes.  

The command `export()` is also from the package "rio".  

First, the object that you want to export is named (`linelist_cleaned`).  

Then, the function `paste0()` creates the name for the new file by combining "AmTiman_linelist_cleaned_" with the current date and finally the extention ".xlsx". Don't be confused - this is equivalen of writing `rio::export(linelist_cleaned, "AmTiman_linelist_cleaned_2019-08-24.xlsx")`, but using `paste0()` and `Sys.Date()` means the current date will always be used automatically.

```{r save_cleaned_data, eval = FALSE}
## OPTIONAL: save your cleaned dataset! 
## put the current date in the name so you know!
rio::export(linelist_cleaned, paste0("AmTiman_linelist_cleaned_", Sys.Date(), ".xlsx"))
```


```{r, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# Save items for use in the next page (Person analyses).
saveRDS(linelist_cleaned, "data/linelist_cleaned.rds")
saveRDS(SYMPTOMS, "data/SYMPTOMS.rds")
saveRDS(LABS, "data/LABS.rds")
saveRDS(first_week, "data/first_week.rds")
saveRDS(obs_start, "data/obs_start.rds")
saveRDS(obs_end, "data/obs_end.rds")




```
