---
date: "2019-08-21T10:38:36+02:00"
title: Person analyses
weight: 12
---



```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# hide all code chunks in the output, but show errors
knitr::opts_chunk$set(echo = TRUE,       # hide all code chunks in output
                      error = TRUE,       # show errors if they appear, but don't stop
                      fig.width = 6*1.25, # Figure width
                      fig.height = 6      # Figure height
                     )
# set default NA to - in output, define figure width/height
options(knitr.kable.NA = "-")

# Installing required packages for this template
required_packages <- c("knitr",       # create output docs
                       "dplyr",       # clean/shape data
                       "forcats",     # clean/shape data
                       "stringr",     # clean text
                       "rio",         # read in data
                       "ggplot2",     # create plots and charts
                       "sitrep",      # MSF field epi functions
                       "linelist",    # Functions for cleaning/standardising data
                       "incidence",   # create epicurves
                       "aweek",       # define epi weeks
                       "epitools",    # 2x2 tables and other epi goodness
                       "epitrix",     # epi helpers and tricks
                       "sf",          # encode spatial vector data
                       "ggspatial")   # plot maps

for (pkg in required_packages) {
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  
  # load packages to this current session 
  library(pkg, character.only = TRUE)
}


# set default text size to 16 for plots
# give classic black/white axes for plots
ggplot2::theme_set(theme_classic(base_size = 18))

# Set the day that defines the beginning of your epiweek.
# you can start the week on any day of the week
# (the ISO standard is to start on Monday) 
aweek::set_week_start("Monday")



date2week("2017-04-20")      #Sys.Date() uses the current date from your computer


reporting_week <- readRDS("data/reporting_week.rds") # Set the reporting week

# Read in the R object linelist_cleaned that is defined and saved 
# in the AJS_Walkthrough_Cleaning script
linelist_cleaned <- readRDS("data/linelist_cleaned.rds")

population_data_region <- readRDS("data/population_data.rds")

SYMPTOMS <- readRDS("data/SYMPTOMS.rds")
LABS <- readRDS("data/LABS.rds")
first_week <- readRDS("data/first_week.rds")
obs_start <- readRDS("data/obs_start.rds")
obs_end <- readRDS("data/obs_end.rds")

```






TODO [general description of person section]

### Demographic Tables:

The first demographic table presents patients by their age group (the table's rows) and their relationship with the case definition (the table's columns). This chunk uses piping (%>%, see reference TODO HERE) to link six different functions and produce a table:  

* The tab_linelist() function creates a frequency and percent table of age_group, with columns differentiating by case_def value  
* The select() function removes an unnecessary column "variable" generated by tab_linelist().  
* The rename() function renames the default column "value" as "Age Group"  
* The rename_redundant() function replaces any column name that has proportion with "%"  
* The augment_redundant() command replaces any column name with n with " cases (n)"  
* The kable() command completes the table, in this case with two digits after each decimal


```{r describe_by_age_group_and_def}
# Describe observations by age_group and case_def
tab_linelist(linelist_cleaned, 
             age_group, strata = case_def, 
             col_total = TRUE, row_total = TRUE) %>%
  select(-variable) %>%
  rename("Age group" = value) %>%
  rename_redundant("%" = proportion) %>%
  augment_redundant(" cases (n)" = " n$") %>%
  kable(digits = 2)
```

You can change the argument strata = to refer to any number of variables (in this next example, sex). To show proportions of the total population of observations, add the argument prop_total specified as TRUE, as below. 

```{r total_props_agegroup_sex}
tab_linelist(linelist_cleaned, 
             age_group, strata = sex,
             col_total = TRUE, row_total = TRUE, prop_total = TRUE) %>% 
  select(-variable) %>%
  rename("Age group" = value) %>%
#  rename_redundant("%" = proportion) %>%
  augment_redundant(" cases (n)" = " n$") %>%
  kable(digits = 2)
```

You can also choose to exclude or edit the **in-line code** that gives a count of the missing cases by sex and age_group.  


### Age pyramids 
To print an age pyramid in your report, use the code below. A few things to note:  

* The **variable for split_by should have two non-missing value options** (e.g. Male or Female, Oui or Non, etc.)  
* The variable names work with or without quotation marks  
* The dashed lines in the bars are the midpoint of the un-stratified age group   
* You can adjust the position of the legend by replacing legend.position = "bottom" with "top", "left", or "right"  
* Read more by searching plot_age_pyramid in the Help tab of the lower-right RStudio pane  

You can make this a pyramid of months by supplying age_group_mon to the age_group argument.  



```{r age_pyramid, warning=FALSE}
# plot age pyramid by sex 
plot_age_pyramid(linelist_cleaned, 
                 age_group = "age_group", 
                 split_by = "sex") + 
  labs(y = "Cases (n)", x = "Age group") + # change axis  labels (nb. x/y flip)
  theme(legend.position = "bottom",     # move legend to bottom
        legend.title = element_blank(), # remove title
        text = element_text(size = 18)  # change text size
       )
```

**To have an age pyramid of patients under 2 by month age groups**, it is best to add a filter() step to the beginning of the code chunk, as below. This step selects for linelist_cleaned observations that meet the specified critera and passes that reduced dataset through the "pipes" to the next function:  (plot_age_pyramid()). If this filter() step is not added, you will see that the largest pyramid bars are of "missing" - these are the patients old enough to not have a months age group.  

Note that the filter step does not drop any observations from the linelist_cleaned object itself. Because the filter is **not** being assigned (`<-`) to over-write linelist_cleaned, this filter is only temporarily applied for the purpose of producing the age pyramid.

```{r age_pyramid_under_two, warning=FALSE}
# plot age pyramid by month groups, for observations under 2 years
filter(linelist_cleaned, age_years <= 2) %>%
  plot_age_pyramid(age_group = "age_group_mon",
                   split_by = "sex") +
                  # stack_by = "case_def") +
  labs(y = "Cases (n)", x = "Age group (months)") + # change axis  labels (nb. x/y flip)
  theme(legend.position = "bottom",     # move legend to bottom
        legend.title = element_blank(), # remove title
        text = element_text(size = 18)  # change text size
       )
```

### Inpatient statistics  

The text following the age pyramids uses **in-line code** to describe the distribution of outpatient and inpatient observations, and descriptive statistics of the length of stay for inpatients. **The Am Timan dataset does not have length of stay**, so it is best to delete those sentences for the final report.  

### Symptom and lab descriptives  

This next code also uses the tab_linelist() function to **create descriptive tables of all the variables that were included in the SYMPTOMS variable list**.  

* In the tab_linelist() function, SYMPTOMS (the value supplied to the second argument) is an object we defined in data cleaning that is a list of variables to tabulate. If this code produces an error about an "Unknown column", ensure that the variables in the object SYMPTOMS are *all* present in your dataset (and spelled correctly). 
* Also in tab_linelist(), the argument `keep = ` must represent the character value to be counted for the the table. As these Am Timan variables are still in French, we change `keep = "Yes"` to `keep = "Oui"`.  
* The mutate() function is aesthetically changing variable names with underscores to spaces.   

```{r describe_by_symptoms, warning=FALSE}
# get counts and proportions for all variables named in SYMPTOMS
tab_linelist(linelist_cleaned, SYMPTOMS, keep = "Oui") %>% 
  select(-value) %>%
  # fix the way symptom names are displayed
  mutate(variable = str_to_sentence(gsub("_", " ", variable))) %>%
  # rename accordingly
  rename_redundant("%" = proportion) %>%
  augment_redundant(" (n)" = " n$") %>%
  kable(digits = 2)
```

The code for the lab table is very similar, but has these differences:  

* TODO  
* TODO  
* TODO  

TODO: Need to fix in data exploration/cleaning phase the lab values (pos, Pos, weakly pos, etc.)
```{r describe_by_labs, warning=FALSE}
# get counts and proportions for all variables named in LABS
tab_linelist(linelist_cleaned, LABS, 
             transpose = "value") %>% 
  # fix the way lab test names are displayed
  mutate(variable = str_to_sentence(gsub("_", " ", variable))) %>%
  # rename accordingly
  rename("Lab test" = variable) %>%
  rename_redundant("%" = proportion) %>%
  augment_redundant(" (n)" = " n$") %>%
  kable(digits = 2)
```

## Case Fatality Ratio (CFR)

The opening text with in-line code must be edited to match our Am Timan data. The second in-line code references the variable exit_status and the value "Dead" - **we know that it must now reference the variable exit_status2 and the value "Died"**. Also, we do not have Dead on Arrival recorded in our dataset. Likewise, **the section on time-to-death does not apply to our dataset and can be deleted.**  

TODO: PNG image of in-line code edited to be correct  

Overall CFR is produced with the code below. Note the following:  

* This code requires the variable patient_facility_type, that it has a value "Inpatient", and the DIED variable.  
* A filter is applied that restricts this table to Inpatient observations only.  
* rename() is used to change the column labels in the table  

```{r overall_cfr, warning = FALSE, message = FALSE}
# use arguments from above to produce overal CFR
overall_cfr <- linelist_cleaned %>% 
  filter(patient_facility_type == "Inpatient") %>%
  case_fatality_rate_df(deaths = DIED, mergeCI = TRUE) %>%
  rename("Deaths" = deaths,
         "Cases" = population,
         "CFR (%)" = cfr,
         "95%CI" = ci)
  knitr::kable(overall_cfr, digits = 2)         # print nicely with 2 digits
```

The next code adds arguments to case_fatality_rate_df() such as `group = sex` (which stratified the CFR by sex), and `add_total = TRUE` (which makes a total row across the sex groups).  

```{r cfr_by_sex, warning = FALSE, message = FALSE}
linelist_cleaned %>%
  filter(patient_facility_type == "Inpatient") %>%
  mutate(sex = forcats::fct_explicit_na(sex, "-")) %>%
  case_fatality_rate_df(deaths = DIED, group = sex, mergeCI = TRUE, add_total = TRUE) %>%
  rename("Sex" = sex, 
         "Deaths" = deaths, 
         "Cases" = population, 
         "CFR (%)" = cfr, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2)
```

When creating a table of CFR by age groups, one additional step, using the function complete() is required to ensure that all age_group levels are shown even if they have no observations.  


```{r cfr_by_age_group, warning = FALSE, message = FALSE}
linelist_cleaned %>%
  filter(patient_facility_type == "Inpatient") %>%
  case_fatality_rate_df(deaths = DIED, group = age_group, mergeCI = TRUE, add_total = TRUE) %>%
  tidyr::complete(age_group, 
                  fill = list(deaths = 0, 
                              population = 0, 
                              cfr = 0, 
                              ci = 0)) %>% # Ensure all levels are represented
  rename("Age group" = age_group, 
         "Deaths" = deaths, 
         "Cases" = population, 
         "CFR (%)" = cfr, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2)
```


The commented code below examines **CFR by case definition**. Note that this is dependant upon our working case_def variable.

```{r cfr_by_case_def, warning = FALSE, message = FALSE}
# Use if you have enough confirmed cases for comparative analysis 

 linelist_cleaned %>%
   filter(patient_facility_type == "Inpatient") %>%
   case_fatality_rate_df(deaths = DIED, group = case_def, mergeCI = TRUE, add_total = TRUE) %>%
   rename("Case definition" = case_def, 
          "Deaths" = deaths, 
          "Cases" = population, 
          "CFR (%)" = cfr, 
          "95%CI" = ci) %>% 
   knitr::kable(digits = 2)
```

## Attack Rate
To use the attack rate section, we need to modify the first code slightly. An object "population" is created from the sum of population counts in the population figures. Because we only imported region-based population counts, we must change this command to reflect that we do not have population_data_age, but rather population_data_region

```{r eval=FALSE}
# OLD define population 
# population <- sum(population_data_age$population)
```

Running the correct command and printing the value of population, we see that the sum population across regions is estimated to be 62336.
```{r}
# Correct command for this exercise
population <- sum(population_data_region$population)

population
```

Below gives the attack rate per 10,000 population (N = `r format(population, big.mark = ",")`)
```{r attack_rate}
# calculate the ar
# store as AR to be able to use output for automating text below (inline functions)
ar <- attack_rate(nrow(linelist_cleaned), population, multiplier = 10000)

ar %>%
  merge_ci_df(e = 3) %>% # merge the lower and upper CI into one column
  rename("Cases (n)" = cases, 
         "Population" = population, 
         "AR (per 100,000)" = ar, 
         "95%CI" = ci) %>% 
  select(-Population) %>% # drop the population column as it is not changing
  knitr::kable(digits = 2, align = "r")
```

We are unable to calculate the attack rate by age group, because we do not have population counts for each age group.

Mortality attributable to AJS - not appropriate for this example because...


```{r, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# Save linelist_cleaned for use in the next page (time analyses).
saveRDS(linelist_cleaned, "data/linelist_person_cleaned.rds")
```


