---
date: "2019-08-21T11:00:29+02:00"
title: Time analyses
weight: 13
---




```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# hide all code chunks in the output, but show errors
knitr::opts_chunk$set(echo = TRUE,       # hide all code chunks in output
                      error = TRUE,       # show errors if they appear, but don't stop
                      fig.width = 6*1.25, # Figure width
                      fig.height = 6      # Figure height
                     )
# set default NA to - in output, define figure width/height
options(knitr.kable.NA = "-")

# Installing required packages for this template
required_packages <- c("knitr",       # create output docs
                       "dplyr",       # clean/shape data
                       "forcats",     # clean/shape data
                       "stringr",     # clean text
                       "rio",         # read in data
                       "ggplot2",     # create plots and charts
                       "sitrep",      # MSF field epi functions
                       "linelist",    # Functions for cleaning/standardising data
                       "incidence",   # create epicurves
                       "aweek",       # define epi weeks
                       "epitools",    # 2x2 tables and other epi goodness
                       "epitrix",     # epi helpers and tricks
                       "sf",          # encode spatial vector data
                       "ggspatial")   # plot maps

for (pkg in required_packages) {
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  
  # load packages to this current session 
  library(pkg, character.only = TRUE)
}


# set default text size to 16 for plots
# give classic black/white axes for plots
ggplot2::theme_set(theme_classic(base_size = 18))

# Set the day that defines the beginning of your epiweek.
# you can start the week on any day of the week
# (the ISO standard is to start on Monday) 
aweek::set_week_start("Monday")



date2week("2017-04-20")      #Sys.Date() uses the current date from your computer


reporting_week <- readRDS("data/reporting_week.rds") # Set the reporting week

# Read in the R objects that were defined in previous pages
linelist_cleaned <- readRDS("data/linelist_person_cleaned.rds")

population_data_region <- readRDS("data/population_data.rds")

SYMPTOMS <- readRDS("data/SYMPTOMS.rds")
LABS <- readRDS("data/LABS.rds")
first_week <- readRDS("data/first_week.rds")
obs_start <- readRDS("data/obs_start.rds")
obs_end <- readRDS("data/obs_end.rds")

```


TODO: Introduction to time analyses...  



The first code chunk (create_incidence) is about setting-up the data for the epidemic curves. There will not be output from this first code chunk.  

**First**, the function incidence() counts the number of cases for each week, by date_of_onset. It stores this information in the object inc_week_7.  

**Second**, the theme elements are set that ggplot2() will use when plotting the epidemic curves (you can read more about ggplot2 and theme elements HERE TODO). This includes the angle and placement of x-axis label text, having no title in the legend, and setting colors and type for x and y gridlines.  

**Third**, the text labels for the x and y axes, title, and subtitle (subtitle automatically using the reporting week defined at the top of the template). These are stored in the object epicurve_labels.  


```{r create_incidence, message = FALSE}
# This code creates case counts for each week of your outbreak, overall
# As with aweek, you can change the start of your week to e.g. "Sunday week"
inc_week_7 <- incidence(linelist_cleaned$date_of_onset, interval = "Monday week")

# this sets the theme in ggplot for epicurves
epicurve_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
  legend.title = element_blank(),
  panel.grid.major.x = element_line(color = "grey60", linetype = 3),
  panel.grid.major.y = element_line(color = "grey60", linetype = 3)
)

# This sets the labels in ggplot for the epicurves
epicurve_labels <- labs(x = "Calendar week", 
                        y = "Cases (n)", 
                        title = "Cases by week of onset",
                        subtitle = sprintf("Source: MSF data from %s", reporting_week)
                       ) 
```

In this next chunk (epicurve), the plot of the epidemic curve is finalized and displayed. note that the plot() function from the package incidence(), that is used here, uses ggplot() and is different than the base R plot() function.  

As with ggplot2, layers of the graphic are added using the + symbol. In this case, the plot from incidence() is stored first, then the axes, then the labels, and finally the theme parameters. All of these are stored in the object basic_curve, which is finally run to display the result.  

```{r epicurve, message = FALSE}
# plot your epicurve as a ggplot (incidence::plot is different to base::plot)
basic_curve <- plot(inc_week_7, show_cases = TRUE, border = "black", n_breaks = nrow(inc_week_7)) + 
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # add labels to axes and below chart
  epicurve_labels +
  # change visuals of dates and remove legend title
  epicurve_theme
# show your plot (stored for later use) 
basic_curve
## if the outbreak has been going on for a while, your x-axis might look messy.
# to reduce the number of labels, uncomment the below.
# you can customize the number of breaks by changing n_breaks.
#
#basic_curve + scale_x_incidence(inc_week_7, n_breaks = 6)
```

The x-axis is quite crowded, so we can uncomment and use the helper code at the bottom of the code chunk. This modifies the already-defined basic_curve (using a +) to change the x-axis to show a specified number of dates (6 in this case).  

```{r}
# Modifies basic_curve to show only 6 x-axis breaks 
basic_curve + scale_x_incidence(inc_week_7, n_breaks = 6)
```

Other possible variants on the epi curve:  
* Option 1  
* Option 2  
* Option 3  


The below gives the attack rate per week. 

```{r attack_rate_per_week, warning = FALSE, message = FALSE}
# counts and cumulative counts by week
cases <- linelist_cleaned %>%
  arrange(date_of_onset) %>%        # arrange by date of onset
  count(epiweek, .drop = FALSE) %>% # count all epiweeks and include zero counts
  mutate(cumulative = cumsum(n))    # add a cumulative sum

# attack rate for each week
ar <- attack_rate(cases$n, population, multiplier = 100000) %>% 
  bind_cols(select(cases, epiweek), .) # add the epiweek column to table

ar %>%
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Cases (n)" = cases, 
         "Population" = population, 
         "AR (per 100,000)" = ar, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```

And cumulatively:  


```{r cumulative_attack_rate_per_week}
# cumulative attack rate by week
attack_rate(cases$cumulative, population, multiplier = 100000) %>% 
  bind_cols(select(cases, epiweek), .) %>% # add the epiweek column to table
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Cases (n)" = cases, 
         "Population" = population, 
         "AR (per 100,000)" = ar, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```

```{r cfr_per_week, warning = FALSE, message = FALSE}
# group by known outcome and case definition 
cfr <- linelist_cleaned %>%
  filter(patient_facility_type == "Inpatient") %>%
  case_fatality_rate_df(grepl("Dead", exit_status), group = epiweek)

cfr %>%
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Deaths" = deaths, 
         "Cases" = population, 
         "CFR (%)" = cfr, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```

You could plot the AR (in the population) and CFR (among inpatients only) together as line graphs by epiweek. 

```{r ar_line_graph}
ar_plot <- ggplot(ar, aes(x = week2date(epiweek) + (7 * 0.5), group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              color = "blue", fill = "blue", linetype = 2, alpha = 0.2, show.legend = FALSE) +
  geom_line(aes(y = ar), color = "blue", show.legend = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # scale the x axis the same as the incidence curve. Expand forces it to align. 
  incidence::scale_x_incidence(inc_week_7, n_breaks = nrow(inc_week_7), expand = c(0, 7 * 1.5)) +
  # add labels to axes and below chart
  labs(x = "Calendar week", y = "AR [95% CI]", subtitle = "Attack Rate (per 100,000)") + 
  # change visuals of dates and remove legend title
  epicurve_theme
```

To plot CFR you must first make the CFR table using above code

```{r cfr_line_graph}
cfr_plot <- ggplot(cfr, aes(x = week2date(epiweek) + (7 * 0.5), group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              color = "red", fill = "red", linetype = 2, alpha = 0.2, show.legend = FALSE) +
  geom_line(aes(y = cfr), color = "red", show.legend = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # scale the x axis the same as the incidence curve. Expand forces it to align. 
  incidence::scale_x_incidence(inc_week_7, n_breaks = nrow(inc_week_7), expand = c(0, 7 * 1.5)) +
  # add labels to axes and below chart
  labs(x = "Calendar week", y = "CFR [95% CI]", 
       subtitle = "Case Fatality Ratio [95% CI] Among Inpatients") + 
  # change visuals of dates and remove legend title
  epicurve_theme 

```


You could then also add the AR and CFR by week on to an epicurve. 

```{r epicurve_ar_cfr, message = FALSE, fig.height = 10}
nofx <- theme(axis.text.x = element_blank(),
              axis.title.x = element_blank())
cowplot::plot_grid(
  basic_curve + nofx,
  ar_plot + nofx,
  cfr_plot,
  align = "v", # align plots vertically
  axis = "lr", # only by their left and right margins
  ncol = 1     # allow only one column
)
```


more optins...



```{r, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# Save linelist_cleaned for use in the next page (time analyses).
saveRDS(linelist_cleaned, "data/linelist_time_cleaned.rds")
```


