---
date: "2019-08-21T11:00:29+02:00"
title: Time analyses
weight: 13
---




```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# hide all code chunks in the output, but show errors
knitr::opts_chunk$set(echo = TRUE,       # hide all code chunks in output
                      error = TRUE,       # show errors if they appear, but don't stop
                      fig.width = 6*1.25, # Figure width
                      fig.height = 6      # Figure height
                     )
# set default NA to - in output, define figure width/height
options(knitr.kable.NA = "-")

# Installing required packages for this template
required_packages <- c("knitr",       # create output docs
                       "dplyr",       # clean/shape data
                       "forcats",     # clean/shape data
                       "stringr",     # clean text
                       "rio",         # read in data
                       "ggplot2",     # create plots and charts
                       "sitrep",      # MSF field epi functions
                       "linelist",    # Functions for cleaning/standardising data
                       "incidence",   # create epicurves
                       "aweek",       # define epi weeks
                       "epitools",    # 2x2 tables and other epi goodness
                       "epitrix",     # epi helpers and tricks
                       "sf",          # encode spatial vector data
                       "ggspatial",
                       "here")   # plot maps

for (pkg in required_packages) {
  # install packages if not already present
  if (!pkg %in% rownames(installed.packages())) {
    install.packages(pkg)
  }
  
  # load packages to this current session 
  library(pkg, character.only = TRUE)
}


# set default text size to 16 for plots
# give classic black/white axes for plots
ggplot2::theme_set(theme_classic(base_size = 18))

# Set the day that defines the beginning of your epiweek.
# you can start the week on any day of the week
# (the ISO standard is to start on Monday) 
aweek::set_week_start("Monday")



date2week("2017-04-20")      #Sys.Date() uses the current date from your computer


reporting_week <- readRDS("data/reporting_week.rds") # Set the reporting week

# Read in the R objects that were defined in previous pages
linelist_cleaned <- readRDS("data/linelist_person_cleaned.rds")

population_data_region <- readRDS("data/population_data.rds")
population <- readRDS("data/population.rds")

SYMPTOMS <- readRDS("data/SYMPTOMS.rds")
LABS <- readRDS("data/LABS.rds")
first_week <- readRDS("data/first_week.rds")
obs_start <- readRDS("data/obs_start.rds")
obs_end <- readRDS("data/obs_end.rds")

```


TODO: Introduction to time analyses...  


### Basic epidemic curve  

The first code chunk (create_incidence) is about setting-up the data for the epidemic curves. *There will not be output from this first code chunk.*  

**First**, the function incidence() **counts the number of cases** for each week, by date_of_onset. It stores this information in the object inc_week_7. Read more about incidence() in the Help pane. 

**Second**, the **theme elements** are set that ggplot2() will use when plotting the epidemic curves (you can read more about ggplot2 and theme elements HERE TODO). This includes the angle and placement of x-axis label text, having no title in the legend, and setting colors and type for x and y gridlines.  

**Third**, the **text labels** for the x and y axes, title, and subtitle (subtitle automatically using the reporting week defined at the top of the template). These are stored in the object epicurve_labels.  

**Note: It is necessary to use the "full name" of variables for this to work correctly - this means date_of_onset should be listed in incidence() as `linelist_cleaned$date_of_onset`.**


```{r create_incidence, message = FALSE}
# This code creates case counts for each week of your outbreak, overall
# As with aweek, you can change the start of your week to e.g. "Sunday week"
inc_week_7 <- incidence(linelist_cleaned$date_of_onset, interval = "Monday week")

# this sets the theme in ggplot for epicurves
epicurve_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), 
  legend.title = element_blank(),
  panel.grid.major.x = element_line(color = "grey60", linetype = 3),
  panel.grid.major.y = element_line(color = "grey60", linetype = 3)
)

# This sets the labels in ggplot for the epicurves
epicurve_labels <- labs(x = "Calendar week", 
                        y = "Cases (n)", 
                        title = "Cases by week of onset",
                        subtitle = sprintf("Source: MSF data from %s", reporting_week)
                       ) 
```

**In this next chunk (epicurve), the plot of the epidemic curve is finalized and displayed.**  

*Note that the plot() function used here is from the package incidence(), not the base R plot() function. This one uses ggplot2() and so has different syntax and structure. As with ggplot2, layers of the graphic are added sequentially using the + symbol at the end of each sub-command.*  

In the example below, the weekly count data from inc_week_7 are plotted first, then miscellaneous settings are specified: the y-axis origin is set, the labels, and finally the theme parameters. All of these elements are stored in the object "basic_curve", which is finally run to display the plot.

**Note: It is necessary to use the "full name" of variables for this to work correctly - this means date_of_onset should be listed in incidence() as `linelist_cleaned$date_of_onset`.** Otherwise you may get an error such as "object 'date_of_onset' not found". 

```{r epicurve, message = FALSE}
# plot your epicurve as a ggplot (incidence::plot is different to base::plot)
basic_curve <- plot(inc_week_7, show_cases = TRUE, border = "black", n_breaks = nrow(inc_week_7)) + 
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # add labels to axes and below chart
  epicurve_labels +
  # change visuals of dates and remove legend title
  epicurve_theme
# show your plot (stored for later use) 
basic_curve

```

This x-axis is difficult to read with every week displayed. To increase read-ability we can uncomment and use the helper code at the bottom of that code chunk. This modifies the already-defined "basic_curve" object (with a +) to specify that the x-axis shows only a specified number of breaks/dates (in this case 6). Note: A warning that a scale for x is being replaced is okay.  

*Note that the data are still shown by week - it is only the labels on the scale that have changed.*   

```{r, warning=FALSE, message=FALSE}
# Modifies basic_curve to show only 6 breaks in the x-axis 
basic_curve + scale_x_incidence(inc_week_7, n_breaks = 6)
```
**To change the intervals used for the data aggregation themselves**, use the `interval =` argument in the original incidence() command. There are the options to do this biweekly, by month, quarter or year (search the Help pane for the incidence() function in the incidence package). **Below is a biweekly example** (the axes labels have been returned to default - showing each two-week increment).  

*Note: In the code below, the plot is not saved to an object and simply prints when the plot() command and it's various added (+) elements such as scale, labels, and theme are run.*

```{r biweekly_epicurve, message = FALSE}
# interval is set to "2 Monday weeks" for biweekly case aggregation
inc_week_14 <- incidence(linelist_cleaned$date_of_onset, interval = "2 Monday weeks")

# plot the epicurve
plot(inc_week_14, show_cases = TRUE, border = "black", n_breaks = nrow(inc_week_14)) + 
  scale_y_continuous(expand = c(0,0)) +  # set origin for axes
  # add labels to axes and below chart
  epicurve_labels +
  # change visuals of dates and remove legend title
  epicurve_theme
```

To **stratify the epidemic curve** by a categorical variable (such as gender), the `groups = ` argument is added to the incidence() function. Additionally, the `show_cases = ` argument of plot() can be set to FALSE to improve aesthetics.  

**Note: Remember that if listing a variable, the plot() function expects its "full name", such as `linelist_cleaned$sex`**. Otherwise you may get an error such as "object 'sex' not found".  

```{r incidence_by_gender, message = FALSE}
# get counts by gender
inc_week_7_gender <- incidence(linelist_cleaned$date_of_onset, 
                               interval = "2 Monday weeks", 
                               groups = linelist_cleaned$sex)
# plot your epicurve
# here we remove the boxes around each case as it makes gender colours hard to see! (show_cases = FALSE)
plot(inc_week_7_gender, show_cases = FALSE, border = "black", n_breaks = nrow(inc_week_7_gender)) + 
  scale_y_continuous(expand = c(0,0)) +  # set origin for axes
  # add labels to axes and below chart
  epicurve_labels +
  # change visuals of dates, remove legend title and move legend to bottom
  epicurve_theme
```

Lastly, we can **limit the entire plot to a subset of observations** by first filtering the dataset and passing the result to the incidence() function.  

In the example below, incidence() operates upon a filtered version of linelist_cleaned that only includes inpatients. The resulting object is used, as before, in plot() in the step below.  


```{r incidence_by_sex_facility, message = FALSE}
# filter the dataset and pass it to the incidence() function
inc_week_7_sex_fac <- linelist_cleaned %>%
  filter(patient_facility_type == "Inpatient") %>%
  with(incidence(date_of_onset, interval = "2 Monday weeks", groups = sex))

# plot as before
plot(inc_week_7_sex_fac, show_cases = TRUE, border = "black", n_breaks = nrow(inc_week_7_sex_fac)) + 
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # add labels to axes and below chart
  epicurve_labels +
  # change visuals of dates, remove legend title and move legend to bottom
  epicurve_theme
```


### Analysis of other indicators by time  


#### Attack rate by week  

The code below produces a table of the attack rates by epiweek.  

**First**, the object "cases" is defined as linelist observations counted and arranged/ordered by epiweek, accompanied by a cumulative sum column.  

**Second**, the object "ar" is defined using the attack_rate() function, which looks with "cases" at counts for each week (column "n") and the denominator "population" which was defined earlier as the sum of the regional populations (62336). The rate multiplier is set to 10000 but is changeable. The epiweeks are also attached to "ar" as a column.  

**Third**, aesthetic operations are done to "ar" to improve column names. Finally, the table is produced.  


```{r attack_rate_per_week, warning = FALSE, message = FALSE}
# counts and cumulative counts by week
cases <- linelist_cleaned %>%
  arrange(date_of_onset) %>%        # arrange by date of onset
  count(epiweek, .drop = FALSE) %>% # count all epiweeks and include zero counts
  mutate(cumulative = cumsum(n))    # add a cumulative sum

# attack rate for each week
ar <- attack_rate(cases$n, population, multiplier = 10000) %>% 
  bind_cols(select(cases, epiweek), .) # add the epiweek column to table

# Display the table
ar %>%
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Cases (n)" = cases, 
         "Population" = population, 
         "AR (per 10,000)" = ar, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```
 

This code produces the *cumulative* attack rate, by week. In the attack_rate() command it references the `cases$cumulative` column, instead of `cases$n column`.


```{r cumulative_attack_rate_per_week}
# cumulative attack rate by week
attack_rate(cases$cumulative, population, multiplier = 10000) %>% 
  bind_cols(select(cases, epiweek), .) %>% # add the epiweek column to table
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Cases (n)" = cases, 
         "Population" = population, 
         "AR (per 10,000)" = ar, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```

This next code chunk applies a filter for only inpatient observations. It also uses a different function, case_fatality_rate_df(), to produce a table showing deaths, cases, CFR, and CFR confidence intervals.  

```{r cfr_per_week, warning = FALSE, message = FALSE}
# group by known outcome and case definition 
cfr <- linelist_cleaned %>%
  filter(patient_facility_type == "Inpatient") %>%
  case_fatality_rate_df(grepl("Dead", exit_status), group = epiweek)


cfr %>%
  merge_ci_df(e = 4) %>% # merge the lower and upper CI into one column
  rename("Epiweek" = epiweek, 
         "Deaths" = deaths, 
         "Cases" = population, 
         "CFR (%)" = cfr, 
         "95%CI" = ci) %>% 
  knitr::kable(digits = 2, align = "r")
```

#### Plotting AR, CFR, and Epidemic Curve together  

These code chunks produce plots of the AR (in the population), CFR (among inpatients only), and the AF and CFR as line graphs together with the Epidemic Curve.  

**The first chunk prepares a plot of attack rate by epiweek, but does not yet print it.** You can print it alone by running a command of its name `ar_plot`.

*Note: this ar_plot depends on you having already run the attack_rate_per_week chunk. If you have not, you may see "Error: epiweek not found"*  

```{r ar_line_graph}
ar_plot <- ggplot(ar, aes(x = week2date(epiweek) + (7 * 0.5), group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              color = "blue", fill = "blue", linetype = 2, alpha = 0.2, show.legend = FALSE) +
  geom_line(aes(y = ar), color = "blue", show.legend = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # scale the x axis the same as the incidence curve. Expand forces it to align. 
  incidence::scale_x_incidence(inc_week_7, n_breaks = nrow(inc_week_7), expand = c(0, 7 * 1.5)) +
  # add labels to axes and below chart
  labs(x = "Calendar week", y = "AR [95% CI]", subtitle = "Attack Rate (per 10,000)") + 
  # change visuals of dates and remove legend title
  epicurve_theme
```

**The second code chunk prepares a plot of CFR by epiweek, but does not yet print it.** You can print it alone by running a command with its name `cfr_plot`.  

*Note: this cfr_plot depends on you having already run the cfr_per_week chunk. If you have not, you may see "Error: epiweek not found".*  

```{r cfr_line_graph}
cfr_plot <- ggplot(cfr, aes(x = week2date(epiweek) + (7 * 0.5), group = 1)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              color = "red", fill = "red", linetype = 2, alpha = 0.2, show.legend = FALSE) +
  geom_line(aes(y = cfr), color = "red", show.legend = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +  # set origin for axes
  # scale the x axis the same as the incidence curve. Expand forces it to align. 
  incidence::scale_x_incidence(inc_week_7, n_breaks = nrow(inc_week_7), expand = c(0, 7 * 1.5)) +
  # add labels to axes and below chart
  labs(x = "Calendar week", y = "CFR [95% CI]", 
       subtitle = "Case Fatality Ratio [95% CI] Among Inpatients") + 
  # change visuals of dates and remove legend title
  epicurve_theme 
```


**Adding the AR and CFR plots with an Epidemic Curve** and plotting them all together is now possible with the function cowplot(), which nicely combines different plots. 

```{r epicurve_ar_cfr, message = FALSE, fig.height = 10, warning=FALSE}
# Define no specified axis text or title over the combined plot
nofx <- theme(axis.text.x = element_blank(),
              axis.title.x = element_blank())

# Combine the three plots together
cowplot::plot_grid(
  basic_curve + nofx,
  ar_plot + nofx,
  cfr_plot,
  align = "v", # align plots vertically
  axis = "lr", # only by their left and right margins
  ncol = 1     # allow only one column
)
```


#### Admission and exits by epiweek  

This next code chunk describes each epiweek's admission by their case definition status. It uses tab_linelist() again to look at linelist_cleaned, with rows as epiweek, and columns as case_def, with row and column totals.  


```{r describe_admissions_by_epiweek, warning = FALSE}
# get counts and props of admissions by epiweek and case definition 
# include column and row totals 
tab_linelist(linelist_cleaned, epiweek, strata = case_def,
             col_total = TRUE,
             row_total = TRUE) %>% 
  select(-variable) %>%
  rename_redundant("%" = proportion) %>%
  augment_redundant(" (n)" = " n$") %>%
  kable(digits = 2)
```

Finally, this produces the same table as above but uses exit_status2 as the strata/column variable to display exits per epiweek. Note that this is also filtered to only include inpatient observations.  


```{r describe_exits_by_epiweek, warning = FALSE}
# get counts and props of admissions by epiweek among inpatients
# include column and row totals 
linelist_cleaned %>% 
  filter(patient_facility_type == "Inpatient") %>%
  tab_linelist(epiweek, strata = exit_status2, 
               col_total = TRUE, row_total = TRUE) %>% 
  select(-variable) %>%
  rename("Week" = value) %>%
  rename_redundant("%" = proportion) %>%
  augment_redundant(" (n)" = " n$") %>%
  kable(digits = 2)
```




```{r, include=FALSE, results='hide', message=FALSE, warning=FALSE}
# Save linelist_cleaned for use in the next page (time analyses).
saveRDS(linelist_cleaned, "data/linelist_time_cleaned.rds")
```


